### 2015-08-14 Manuscript/SW_early_life_telomeres_SOM_v9.Rmd
Added new data, included chicks, updated figures

#Run the model
lm1 <- lmer(TLKB ~ TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
            data = FlSAall,REML=FALSE)

@@

```{r Figure S3,fig.width=5}

temp <- subset(FlSA,LayYear %in% c(1999,2006,2007,2000))

FigS3A <- ggplot(temp,
      aes(x=GroupSize,
      y=TLKB,
      col=LayYear)) +
  stat_smooth(method=lm,fullrange=T,alpha=0.2, aes(fill=LayYear),lwd=1.2) + 
    ylab('Telomere length (kb)') + 
  xlab('Local Density') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  scale_color_brewer(palette='Set1') + 
  scale_fill_brewer(palette='Set1') +
  annotate('text',x=2,y=10,label='A')

FigS3B <- ggplot(temp,
      aes(x=Condition,
      y=TLKB,
      col=LayYear)) +
  stat_smooth(method=lm,fullrange=T,alpha=0.2, aes(fill=LayYear),lwd=1.2) + 
    ylab('Telomere length (kb)') + 
  xlab('Body Condition') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  scale_color_brewer(palette='Set1') + 
  scale_fill_brewer(palette='Set1') +
  annotate('text',x=-2,y=15,label='B')

multiplot(FigS3A,FigS3B)

**Figure S3** Within cohort relationships between telomere length and **A** local density and **B** body condition in juvenile Seychelles warblers. Only years with juvenile sample sizes >20 are included. Lines and shaded areas represent fitted values and 95% confidence limits from linear regressions.

>>>>

#Run the model
lm1 <- lmer(TLKB ~ Agemonths + TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
            data = FlSAall,REML=FALSE)

@@

### 2015-08-14 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
Added new data, included chicks, updated figures

#Run the model
lm1 <- lmer(TLKB~ TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
            data = FlSAall,REML=FALSE)

@@

#Sort out row/col names
nams <- rownames(table1)
nams[nams == 'c.Ageclass'] <- 'Age class'
nams[nams == 'z.Condition'] <- 'Body condition'
nams[nams == 'z.GroupSize'] <- 'Local density'
nams[nams == 'c.Sex'] <- 'Sex'
nams[nams == 'z.Insect'] <- 'Food av.'
nams[nams == 'z.TQspace'] <- 'TQ'
nams[nams == 'z.Helper'] <- 'Helpers'

@@

ddFig1B <- ddply(temp,
                 .(Insect),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = 1.96*se(TLKB))
ddFig1B <- ddFig1B[complete.cases(ddFig1B),]
ddFig1B$Insect <- as.numeric(paste(ddFig1B$Insect))

Fig1B <- ggplot(ddFig1B,
       aes(x = Insect,
           y = TLKBmean)) +
  ylab('Telomere length (kb)') + 
  xlab('Annual food availability') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  stat_smooth(data=FlSAall,aes(x=Insect,y=TLKB),method=lm,col='black',se=F) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
  scale_colour_manual(values = wes_palette("Royal1")) +
  annotate('text',x=8,y=12,label='B')

Fig1C <- ggplot(FlSAall,
                aes(x = Condition,
                    y = TLKB)) +
    geom_point() +
  ylab('Telomere length (kb)') + 
  xlab('Body Condition') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  stat_smooth(method=lm,col='black',se=F) +
  scale_colour_manual(values = wes_palette("Royal1")) +
  annotate('text',x=-2.5,y=15,label='C')

@@

@@

```{r Figure 2}

x <- summary(lm(TLKB~Insect,data= FlSAall))$resid
xf <- ifelse(x>median(x),'Long','Short')

s1 <- survfit(Surv(time = Lifespan+1, event = Died)~xf, data = FlSAall)
Fig2 <- ggsurv(s1,plot.cens=F) + 
  theme_classic() +
  ylab('Proportion Survived') + 
  xlab('Age (years)') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16),
        legend.position = c(0.8,0.8),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)) +
  scale_colour_manual(name = 'Residual TL',
                      values = c('darkred','red')) +
  guides(linetype=F) +
  theme(legend.title = element_text('Telomere length'))

Fig2

>>>>

#Run the model
lm1 <- lmer(TLKB ~ TQspace + Insect + NonHelper + Sex + Tarsus + BodyMass + Helper + (1|LayYear),
            data = chickall,REML=FALSE)

@@

#Sort out row/col names
nams <- rownames(table1)
nams[nams == 'c.Ageclass'] <- 'Age class'
nams[nams == 'z.BodyMass'] <- 'Body mass'
nams[nams == 'z.Tarsus'] <- 'Tarsus'
nams[nams == 'z.NonHelper'] <- 'Non helpers'
nams[nams == 'c.Sex'] <- 'Sex'
nams[nams == 'z.Insect'] <- 'Food av.'
nams[nams == 'z.TQspace'] <- 'TQ'
nams[nams == 'z.Helper'] <- 'Helpers'

@@

######################################
temp <- chickall
temp$Helper <- ifelse(temp$Helper>0,'Helpers present','Helpers absent')

ddFig1B <- ddply(temp,
                 .(Helper),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = se(TLKB))

Fig1B <- ggplot(ddFig1B,
       aes(x = Helper,
           y = TLKBmean)) +
  ylab('Telomere length (kb)') + 
  xlab('') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
  scale_colour_manual(values = wes_palette("Royal1")) +
  annotate('text',x=0.5,y=7.5,label='B')

Fig1C <- ggplot(chickall,
       aes(x = BodyMass,
           y = TLKB)) +
  ylab('Telomere length (kb)') + 
  xlab('Body mass (g)') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  geom_point() +
  stat_smooth(method = 'lm',col = 'black') +
  annotate('text',x=12,y=15,label='C')

@@

```{r Figure 2,fig.width=5,fig.height=9}
###########
#Fig 2A

#Run the model
lm1 <- lmer(TLKB ~ TQspace + Insect + NonHelper + Sex + Tarsus + BodyMass + Helper + (1|LayYear),
            data = FlSAall,REML=FALSE)

#Sort out row/col names
nams <- rownames(table1)
nams[nams == 'c.Ageclass'] <- 'Age class'
nams[nams == 'z.BodyMass'] <- 'Body mass'
nams[nams == 'z.Tarsus'] <- 'Tarsus'
nams[nams == 'z.NonHelper'] <- 'Non helpers'
nams[nams == 'c.Sex'] <- 'Sex'
nams[nams == 'z.Insect'] <- 'Food av.'
nams[nams == 'z.TQspace'] <- 'TQ'
nams[nams == 'z.Helper'] <- 'Helpers'

Fig2A <- forestplot(table1) +
  annotate('text',x=0.5,y=4,label='A')

ddFig2B <- ddply(temp,
                 .(Insect),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = 1.96*se(TLKB))
ddFig2B <- droplevels(ddFig2B[complete.cases(ddFig2B),])
ddFig2B$Insect <- as.numeric(paste(ddFig2B$Insect))

Fig2B <- ggplot(ddFig2B,
       aes(x = Insect,
           y = TLKBmean)) +
  ylab('Telomere length (kb)') + 
  xlab('Annual food availability') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  stat_smooth(data=FlSAall,aes(x=Insect,y=TLKB),method=lm,col='black',se=T) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
  scale_colour_manual(values = wes_palette("Royal1")) +
  annotate('text',x=2,y=10,label='B')

multiplot(Fig2A,Fig2B)

@@

```{r Figure 3}

s1 <- with(FlSAall,flexsurvreg(Surv(Lifespan+1,Died)~TLKB+Insect, dist="lnorm"))
t1 <- data.frame(s1$res)
t1 <- t1[3:5,]
t1 <- t1[order(rownames(t1)),]
rownames(t1) <- c('Food av.','Tel. length')

Fig2A <- ggplot(t1,aes(x = rownames(t1),y = est)) +
  geom_point() + 
  geom_pointrange(aes(ymin = L95.,ymax = U95.)) +
    theme_classic() +
  ylab('Estimate') + 
  xlab('') + 
  theme_classic() +
  theme(axis.text.x = element_text(angle=90,size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16),
        legend.position = c(0.8,0.8),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)) +
  geom_abline(intercept = 0,slope = 0,lty=2) +
  annotate('text',x=0.6,y=0.15,label='A')

s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSA)
Fig2B <- ggsurv(s1,plot.cens=F) + 
  theme_classic() +
  ylab('Proportion survived') + 
  xlab('Age (years)') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16),
        legend.position = c(0.8,0.8),
        legend.text = element_text(size = 12),
        legend.title = element_blank()) +
  scale_colour_manual(name = 'TL',
                      values = c('darkred','red')) +
  guides(linetype=F) +
  annotate('text',x=0.5,y=1.1,label='B')

multiplot(Fig2A,Fig2B)

### 2015-08-14 Manuscript/SW_early_life_telomeres_v9.Rmd
Added new data, included chicks, updated figures

---
output:
  word_document:
    fig_caption: yes
    highlight: null
bibliography: mybib.bib
csl: ProceedingsB.csl
---
```{r load and clean, include = F}

@@

In Seychelles warblers, oxidative stress is linked to natal territory quality [@VandeCrommenacker2011a], but neither natal territory quality nor natal local density are linked to juvenile or adult survival [@Brouwer2006; @Hammers2013]. In this study we analyse patterns of early-life telomere length and subsequent survival to test the prediction that telomeres fill a missing link between early-life environmental variation and late-life fitness. Specifically we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) the relationship between telomere length and early-life conditions influences short and long-term survival.

@@

We then used parametric survival analysis to examine variation in telomere length in relation to survival, implemented using the 'flexsurv' package in R. The survival time distribution was chosen by comparing the fit (using visual inspection of plots and a comparison of AIC values) of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data. As explanatory variables of survival we included juvenile telomere length, plus the factors identified as being related to telomere length in the previous analysis. Lay year was included as a random effect.

@@

Early-life telomere length differed substantially among birth years (Figure S2), with longer telomeres in years where insect abundance was high (Figure 1B). Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 1A). The best supported model with early-life telomere length as the response variable contained annual food availability and body condition as explanatory variables (R^2^ = `r round2(summary(lm(TLKB~Insect+GroupSize+Condition,data=FlSAall))$r.squared)`). Early-life telomere length  was positively related to body condition (Figure 1C).

To better visualise cohort effects we plotted the relationship between telomere length and both local density and body condition among cohort years (Figure S3). We found that the relationship between local density and telomere length varied among cohorts, from being negative to positive (Figure S3A). A similar, albeit less pronounced, pattern was observed for within-cohort relationships between telomere length and body condition (Figure S3B).

A log-normal survival model best fitted the Seychelles warbler survival data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`). A survival model including telomere length, body condition and annual food availability showed that only telomere length had an on survival (Fig. 2A). Because of the strong correlation between telomere length and

>>>>

---
output:
  word_document:
    fig_caption: yes
    highlight: null
bibliography: MyBib.bib
csl: ProceedingsB.csl
---
```{r load and clean, include = F}

@@

In Seychelles warblers, oxidative stress is linked to natal territory quality [@VandeCrommenacker2011a], but neither natal territory quality nor natal local density are linked to juvenile or adult survival [@Brouwer2006; @Hammers2013]. In this study we analyse patterns of early-life telomere length and subsequent survival to test the prediction that telomeres fill a missing link between early-life environmental variation and late-life fitness. Specifically we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) having longer telomeres in early life is associated with greater short and long-term survival.

@@

We then used parametric survival analysis to examine variation in juvenile telomere length in relation to survival, implemented using the 'flexsurv' package in R. The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values. As explanatory variables of survival we included juvenile telomere length, plus the factors identified as being related to telomere length in the previous analysis. With this approach we aimed to assess the effect of juvenile telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere length.

@@

Early-life telomere length differed substantially among birth years (Figure S2), with longer telomeres in years where insect abundance was high (Figure 1B). Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 1A; Table S1). The best supported model with early-life telomere length as the response variable contained annual food availability and body condition as explanatory variables (Table S1; R^2^ = `r round2(summary(lm(TLKB~Insect+GroupSize+Condition,data=FlSAall))$r.squared)`). Early-life telomere length  was positively related to body condition (Figure 1C).

A log-normal survival model best fitted the Seychelles warbler data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`). A survival model including juvenile telomere length, body condition and early-life food availability showed that only juvenile telomere length had an effect on survival (estimate = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,1])`, CI = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,2])`-`r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB+Insect), dist="lnorm")$res[3,3])`; Fig. 2A). This survival effect was most pronounced in individuals between five and ten years of age (Fig. 2B). Because telomere length and annual food availability are highly correlated (*R* = `r round2(cor(FlSAall$Insect,FlSAall$TLKB)) #confidence test`), we ran an analysis of early-life telomere length on survival excluding early-life food availability, and the telomere effect beacme marginally non-significant (estimate = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,1])`, CI = `r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,2])`-`r round2(flexsurvreg(with(FlSAall,Surv(Lifespan+1,Died)~TLKB), dist="lnorm")$res[3,3])`).

FlSAall$LayYear <- factor(FlSAall$LayYear)
summary(glmer(Lifespan~TLKB+Insect+Condition+(1|LayYear),data=FlSAall,family=poisson))
